```{r include=FALSE}
##Install/Load Packages
ipak <- function(pkg){
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new.pkg))
    install.packages(new.pkg, dependencies = TRUE, repos = "http://cran.us.r-project.org")
  sapply(pkg, require, character.only = TRUE)
}

packages <- c("psych","ggplot2","stargazer","jtools","multcomp","janitor","interactions",
              "tidyverse","scales", "readr", "BayesFactor", "scatterplot3d", "dplyr", "ggstatsplot", "BayesFactor")
ipak(packages)

```
# Exclusion criteria:
email = remove duplicates
missing values on smell taste chemestesis
Sanity check

# Group
Q7__COVID_diagnosis
 
# Descriptive by group
 Q1_1__Year_of_birth                                                                               : num [1:40] 1996 1993 1958 1995 1976 ...
 $ Q2__Country_of_Residence                                                                          : chr [1:40] "USA" "Canada" "USA" "Qatar" ...
 $ Q3__City_of_Residence                                                                             : chr [1:40] "State College" "Montréal, Quėbec" "Falmouth" 
 $ Q4__Gender
"Combustible_cigarette_use_no"
"E-cigarette_use_no"                                                               
 
# Smell
 $ Q12_1__Smell_before_illness                                                                       : num [1:40] 100 79.3 NA NA NA NA 95.8 100 NA 70.8 ...
 $ Q12_2__Smell_during_illness                                                                       : num [1:40] 0.4 11.1 NA NA NA NA 0 7.3 NA 0.9 ...
 $ Q13__1__I_cannot_smell_at_all_/_Smells_smell_less_strong_than_they_did_before                     : num [1:40] 0 1 NA NA NA NA 1 1 NA 0 ...
 $ Q13__2__Smells_smell_different_than_they_did_before_(the_quality_of_smell_has_changed)            : num [1:40] 0 1 NA NA NA NA 0 0 NA 0 ...
 $ Q13__3__I_can_smell_things_that_aren't_there_(for_example_I_smell_burning_when_nothing_is_on_fire): num [1:40] 0 0 NA NA NA NA 1 0 NA 0 ...
 $ Q13__4__Sense_of_smell_fluctuates_(comes_and_goes)
 
# Nasal patency 
 $ Q15_1__Blocked_nose_before_illness                                                                : num [1:40] 0.4 1.2 NA NA NA NA 4.5 0 NA 0.6 ...
 $ Q15_2__Blocked_nose_during_illness    
 
# Taste
 $ Q16_1__Taste_before_illness                                                                       : num [1:40] 91.1 75.5 NA NA NA NA 95.8 100 NA 100 ...
 $ Q16_2__Taste_during_illness                                                                       : num [1:40] 0.7 52.2 NA NA NA NA 3.6 20 NA 0.3 ...
 $ Q17__1__Sweet                                                                                     : num [1:40] 0 0 NA NA NA NA 1 1 NA 0 ...
 $ Q17__2__Salty                                                                                     : num [1:40] 0 0 NA NA NA NA 1 0 NA 0 ...
 $ Q17__3__Sour                                                                                      : num [1:40] 0 0 NA NA NA NA 1 0 NA 0 ...
 $ Q17__4__Bitter                                                                                    : num [1:40] 0 0 NA NA NA NA 1 1 NA 0 ...
 $ Q17__5__Savory/Umami                                                                              : num [1:40] 0 0 NA NA NA NA 0 1 NA 0 ...

#Chemestesis
 $ Q19_1__Chemethesis_before_illness                                                                 : num [1:40] 94.3 33.4 NA NA NA NA 63.5 90 NA 66.3 ...
 $ Q19_2__Chemesthesis_during_illness     
 
```{r import and prepare database}
#Database pulled on April 18th 8:34AM ET by Alyssa Joy Bakke, Penn State University
#Accessed on April 19th 4:00AM ET by Valentina Parma for first check
data <- read_csv("/Users/valentinaparma/Documents/GitHub/GCCR001/analysis/paper/Covid_Results_GCCR001.csv")
str(data)


# Include only the relevant variables for the pre-registered GCCR001 analysis, posted at https://osf.io/auhs8/
keeps <- c("Year_of_birth", "Country_of_Residence", "Gender", "Combustible_cigarette_use_no", "E-cigarette_use_no","Symptoms_changes_in_smell", "COVID_diagnosis", "Smell_before_illness", "Smell_during_illness", "Changes_in_smell_i_cannot_smell_at_all_/_smells_smell_less_strong_than_they_did_before", "Changes_in_smell_smells_smell_different_than_they_did_before_(the_quality_of_smell_has_changed)", "Changes_in_smell_i_can_smell_things_that_aren't_there_(for_example_i_smell_burning_when_nothing_is_on_fire)", "Changes_in_smell_sense_of_smell_fluctuates_(comes_and_goes)", "Blocked_nose_before_illness", "Blocked_nose_during_illness", "Taste_before_illness", "Taste_during_illness", "Changes_in_basic_tastes_sweet", "Changes_in_basic_tastes_salty", "Changes_in_basic_tastes_sour", "Changes_in_basic_tastes_bitter", "Changes_in_basic_tastes_savory/umami", "Chemethesis_before_illness","Chemesthesis_during_illness", "ID")
data_select<-data[keeps]

# Add ID column
# !!!!!!!!Confirm that there are no duplicate IDs that we can identify via emails 
data_select$ID <- seq.int(nrow(data_select))

# Exclusion criteria. Missing values on smell, taste, chemesthesis ratings before and during
data_select_nomiss<-data_select[!is.na(data_select$Smell_before_illness) & 
                                !is.na(data_select$Smell_during_illness) & 
                                !is.na(data_select$Taste_before_illness) & 
                                !is.na(data_select$Taste_during_illness) & 
                                !is.na(data_select$Chemethesis_before_illness) & 
                                !is.na(data_select$Chemesthesis_during_illness), 
                                !is.na(data_select$Blocked_nose_before_illness) & 
                                !is.na(data_select$Blocked_nose_during_illness), ]

# Create the change score
data_select_nomiss$Smell_change<-(data_select_nomiss$Smell_during_illness - data_select_nomiss$Smell_before_illness)
data_select_nomiss$Taste_change<-(data_select_nomiss$Taste_during_illness - data_select_nomiss$Taste_before_illness)
data_select_nomiss$Chemesthesis_change<-(data_select_nomiss$Chemesthesis_during_illness - data_select_nomiss$Chemethesis_before_illness)
data_select_nomiss$Nasal_patency_change<-(data_select_nomiss$Blocked_nose_during_illness - data_select_nomiss$Blocked_nose_before_illness)

# Sanity checks. If "Symptoms changing in smell = yes" then one of these must be yes as well 
#"Changes_in_smell_i_cannot_smell_at_all_/_smells_smell_less_strong_than_they_did_before", #"Changes_in_smell_smells_smell_different_than_they_did_before_(the_quality_of_smell_has_changed)", #"Changes_in_smell_i_can_smell_things_that_aren't_there_(for_example_i_smell_burning_when_nothing_is_on_fire)", #"Changes_in_smell_sense_of_smell_fluctuates_(comes_and_goes)"
# If "Symptoms changing in smell = yes" then the smell change score must be above +/-5

# Simplify the visualization of variables
data_select_nomiss$smell_loss<- data_select_nomiss$`Changes_in_smell_i_cannot_smell_at_all_/_smells_smell_less_strong_than_they_did_before`
data_select_nomiss$parosmia<- data_select_nomiss$`Changes_in_smell_smells_smell_different_than_they_did_before_(the_quality_of_smell_has_changed)`
data_select_nomiss$phantosmia<- data_select_nomiss$`Changes_in_smell_i_can_smell_things_that_aren't_there_(for_example_i_smell_burning_when_nothing_is_on_fire)`
data_select_nomiss$smell_fluctuations<- data_select_nomiss$`Changes_in_smell_sense_of_smell_fluctuates_(comes_and_goes)`
  
# Create a new summary sanity check variable
data_select_nomiss<- as.tibble(data_select_nomiss)
data_select_nomiss$Gender<-recode(data_select_nomiss$Gender, `0` = "Female", `1` = "Male", `2` = "Other", `3` = "Prefer not to say")

data_select_nomiss <- mutate(data_select_nomiss, Sanity_check = ifelse(Symptoms_changes_in_smell == 1 && Smell_change >= 5, 1,
                                             ifelse(Symptoms_changes_in_smell == 1 && Smell_change <= 5, 1,
                                             ifelse(Symptoms_changes_in_smell == 1 && data_select_nomiss$smell_loss == 1, 1,
                                             ifelse(Symptoms_changes_in_smell == 1 && data_select_nomiss$parosmia == 1, 1,
                                             ifelse(Symptoms_changes_in_smell == 1 && data_select_nomiss$phantosmia == 1, 1,
                                             ifelse(Symptoms_changes_in_smell == 1 && data_select_nomiss$smell_fluctuations == 1,1,0)))))))                                             
          

# Create COVID-19 diagnostic groups
# COVID_diagnosis = 1 clinical examination
# COVID_diagnosis = 2 & 3 Lab tested

data_select_nomiss<- mutate(data_select_nomiss, Group = ifelse(COVID_diagnosis == 1, "Clinical exam",
                                                        ifelse(COVID_diagnosis == 2 | COVID_diagnosis == 3, "Lab test", "Remove")))

str(data_select_nomiss)

data_select_nomiss$Age <- 2020-data_select_nomiss$Year_of_birth
data_select_nomiss<-as.data.frame(data_select_nomiss)
keeps2 <- c("Age", "Country_of_Residence", "Gender", "Combustible_cigarette_use_no", "E-cigarette_use_no", "COVID_diagnosis", "Smell_change", "Taste_change", "Chemesthesis_change", "Nasal_patency_change", "smell_loss", "parosmia", "phantosmia", "smell_fluctuations","Changes_in_basic_tastes_sweet", "Changes_in_basic_tastes_salty", "Changes_in_basic_tastes_sour", "Changes_in_basic_tastes_bitter", "Changes_in_basic_tastes_savory/umami", "Sanity_check", "Group", "ID")
data_select_nomiss2<-data_select_nomiss[keeps2]

# Create final database COVID-19 only
data_GCCR001 <- subset(data_select_nomiss2, Group == "Lab test" | Group == "Clinical exam")
table(data_GCCR001$Country_of_Residence)

data_GCCR001$Group<-factor(data_GCCR001$Group)
data_GCCR001$Gender<-factor(data_GCCR001$Gender)
data_GCCR001$Gender <- relevel(data_GCCR001$Gender, ref = "Male")
data_GCCR001$Gender <- relevel(data_GCCR001$Gender, ref = "Female")
str(data_GCCR001)

########################

# Age by Group

# for reproducibility
set.seed(123)
# plot
ggstatsplot::grouped_gghistostats(
  data = data_GCCR001,
  x = Age,
  xlab = "Age (in years))",
  type = "robust", # use robust location measure
  grouping.var = Group, # grouping variable
  title.prefix = "Reported diagnosis",
  normal.curve = TRUE, # superimpose a normal distribution curve
  normal.curve.args = list(color = "red", size = 1),
  ggtheme = ggthemes::theme_tufte(),
  ggplot.component = list( # modify the defaults from `ggstatsplot` for each plot
    ggplot2::scale_x_continuous(breaks = seq(0, 100, 10), limits = (c(0, 100)))
  ),
  messages = FALSE,
  #plotgrid.args = list(nrow = 2),
  title.text = "Age in COVID-19 lab tested vs. clinically examined respondents"
)


# Gender by Group
#!!!! Order of labels and color of the scale need adjustment

tiff("GCCR001_PlotGender.tiff", width = 12, height = 5, units = 'in', res = 300)
# for reproducibility
set.seed(123)

# plot
ggstatsplot::grouped_ggpiestats(
  data = data_GCCR001,
  x = Gender,
  grouping.var = Group, # grouping variable
  title.prefix = "Reported diagnosis", # prefix for the facetted title
  messages = FALSE,
  ggtheme = ggthemes::theme_tufte(),
  title.text = "Gender in COVID-19 lab tested vs. clinically examined respondents"
)

dev.off()


# Smell Change by Group
tiff("GCCR001_PlotSmellChange.tiff", width = 15, height = 5, units = 'in', res = 300)
# for reproducibility
set.seed(123)

# plot
ggstatsplot::grouped_gghistostats(
  data = data_GCCR001,
  x = Smell_change,
  xlab = "Smell change",
  type = "robust", # use robust location measure
  grouping.var = Group, # grouping variable
  title.prefix = "Reported diagnosis",
  normal.curve = TRUE, # superimpose a normal distribution curve
  normal.curve.args = list(color = "red", size = 1),
  ggtheme = ggthemes::theme_tufte(),
  ggplot.component = list( # modify the defaults from `ggstatsplot` for each plot
    ggplot2::scale_x_continuous(breaks = seq(-100, 10, 10), limits = (c(-100, 10)))
  ),
  messages = FALSE,
  #plotgrid.args = list(nrow = 2),
  title.text = "Smell change during - before disease in \n COVID-19 lab tested vs. clinically examined respondents"
)
dev.off()

# Are the distribution similar?
data_GCCR001_Clin<- subset(data_GCCR001, Group == 1)
data_GCCR001_Test<- subset(data_GCCR001, Group == 2)
data_GCCR001_ks<-merge(data_GCCR001_Clin$Smell_change, data_GCCR001_Test$Smell_change)
str(data_GCCR001_ks)
data_GCCR001_ks$Clinical<- data_GCCR001_ks$x
data_GCCR001_ks$Test<- data_GCCR001_ks$y
ks.test(data_GCCR001_ks$Clinical,data_GCCR001_ks$Test)


# Smell change - characterization
tiff("GCCR001_PlotSmellChange_Characterization.tiff", width = 12, height = 5, units = 'in', res = 300)
# for reproducibility
set.seed(123)

# correlation matrix plot
ggstatsplot::grouped_ggcorrmat(
  data = data_GCCR001,
  cor.vars = c(Smell_change, smell_loss, parosmia, phantosmia, smell_fluctuations, Nasal_patency_change),
  cor.vars.names = c(
    "Smell change", # variable names
    "Loss",
    "Parosmia",
    "Phantosmia",
    "Fluctuations",
    "Nasal patency change"
  ),
  type = "robust", # correlation method
  sig.level = 0.001, # threshold of significance
  p.adjust.method = "holm", # p-value adjustment method for multiple comparisons
  colors = c("#B2182B", "white", "#4D4D4D", "red"),
  grouping.var = Group, # grouping variable
  k = 2, # number of digits after decimal point
  title.prefix = "Reported diagnosis",
  messages = FALSE,
  plotgrid.args = list(ncol = 2))
  dev.off()
  
  
# Taste Change by Group
tiff("GCCR001_PlotTasteChange.tiff", width = 15, height = 5, units = 'in', res = 300)
# for reproducibility
set.seed(123)

# plot
ggstatsplot::grouped_gghistostats(
  data = data_GCCR001,
  x = Taste_change,
  xlab = "Taste change",
  type = "robust", # use robust location measure
  grouping.var = Group, # grouping variable
  title.prefix = "Reported diagnosis",
  normal.curve = TRUE, # superimpose a normal distribution curve
  normal.curve.args = list(color = "red", size = 1),
  ggtheme = ggthemes::theme_tufte(),
  ggplot.component = list( # modify the defaults from `ggstatsplot` for each plot
    ggplot2::scale_x_continuous(breaks = seq(-100, 10, 10), limits = (c(-100, 10)))
  ),
  messages = FALSE,
  #plotgrid.args = list(nrow = 2),
  title.text = "Taste change during - before disease in \n COVID-19 lab tested vs. clinically examined respondents"
)
dev.off()


# Are the distribution similar?
data_GCCR001_ks_taste<-merge(data_GCCR001_Clin$Taste_change, data_GCCR001_Test$Taste_change)
str(data_GCCR001_ks_taste)
data_GCCR001_ks_taste$Clinical<- data_GCCR001_ks_taste$x
data_GCCR001_ks_taste$Test<- data_GCCR001_ks_taste$y
ks.test(data_GCCR001_ks_taste$Clinical,data_GCCR001_ks_taste$Test)


# Taste change - characterization
tiff("GCCR001_PlotTasteChange_Characterization.tiff", width = 12, height = 5, units = 'in', res = 300)
# for reproducibility
set.seed(123)

# correlation matrix plot
ggstatsplot::grouped_ggcorrmat(
  data = data_GCCR001,
  cor.vars = c(Taste_change, Changes_in_basic_tastes_sweet, Changes_in_basic_tastes_salty, Changes_in_basic_tastes_sour, Changes_in_basic_tastes_bitter, `Changes_in_basic_tastes_savory/umami`),
  cor.vars.names = c(
    "Taste change", # variable names
    "Sweet",
    "Salt",
    "Sour",
    "Bitter",
    "Umami"
  ),
  type = "robust", # correlation method
  sig.level = 0.001, # threshold of significance
  p.adjust.method = "holm", # p-value adjustment method for multiple comparisons
  colors = c("#B2182B", "white", "#4D4D4D"),
  grouping.var = Group, # grouping variable
  k = 2, # number of digits after decimal point
  title.prefix = "Reported diagnosis",
  messages = FALSE,
  plotgrid.args = list(ncol = 2))
  dev.off()
  
  
# Chemestesis change 
tiff("GCCR001_PlotChemesthesisChange.tiff", width = 15, height = 5, units = 'in', res = 300)
# for reproducibility
set.seed(123)

# plot
ggstatsplot::grouped_gghistostats(
  data = data_GCCR001,
  x = Chemesthesis_change,
  xlab = "Chemesthesis change",
  type = "robust", # use robust location measure
  grouping.var = Group, # grouping variable
  title.prefix = "Reported diagnosis",
  normal.curve = TRUE, # superimpose a normal distribution curve
  normal.curve.args = list(color = "red", size = 1),
  ggtheme = ggthemes::theme_tufte(),
  ggplot.component = list( # modify the defaults from `ggstatsplot` for each plot
    ggplot2::scale_x_continuous(breaks = seq(-100, 100, 10), limits = (c(-100, 100)))
  ),
  messages = FALSE,
  #plotgrid.args = list(nrow = 2),
  title.text = "Chemesthesis change during - before disease in \n COVID-19 lab tested vs. clinically examined respondents"
)
dev.off()


# Are the distribution similar?
data_GCCR001_ks_chemesthesis<-merge(data_GCCR001_Clin$Chemesthesis_change, data_GCCR001_Test$Chemesthesis_change)
str(data_GCCR001_ks_chemesthesis)
data_GCCR001_ks_chemesthesis$Clinical<- data_GCCR001_ks_chemesthesis$x
data_GCCR001_ks_chemesthesis$Test<- data_GCCR001_ks_chemesthesis$y
ks.test(data_GCCR001_ks_chemesthesis$Clinical,data_GCCR001_ks_chemesthesis$Test)


# Nasal patency change 
tiff("GCCR001_PlotNasalPatencyChange.tiff", width = 15, height = 5, units = 'in', res = 300)
# for reproducibility
set.seed(123)

# plot
ggstatsplot::grouped_gghistostats(
  data = data_GCCR001,
  x = Nasal_patency_change,
  xlab = "Nasal patency change",
  type = "robust", # use robust location measure
  grouping.var = Group, # grouping variable
  title.prefix = "Reported diagnosis",
  normal.curve = TRUE, # superimpose a normal distribution curve
  normal.curve.args = list(color = "red", size = 1),
  ggtheme = ggthemes::theme_tufte(),
  ggplot.component = list( # modify the defaults from `ggstatsplot` for each plot
    ggplot2::scale_x_continuous(breaks = seq(-100, 100, 10), limits = (c(-100, 100)))
  ),
  messages = FALSE,
  #plotgrid.args = list(nrow = 2),
  title.text = "Nasal patency change during - before disease in \n COVID-19 lab tested vs. clinically examined respondents"
)
dev.off()
  
# Are the distribution similar?
data_GCCR001_ks_nasalpatency<-merge(data_GCCR001_Clin$Nasal_patency_change, data_GCCR001_Test$Nasal_patency_change)
str(data_GCCR001_ks_nasalpatency)
data_GCCR001_ks_nasalpatency$Clinical<- data_GCCR001_ks_nasalpatency$x
data_GCCR001_ks_nasalpatency$Test<- data_GCCR001_ks_nasalpatency$y
ks.test(data_GCCR001_ks_nasalpatency$Clinical,data_GCCR001_ks_nasalpatency$Test)

```


```{r ratings}
#SMELL
data_GCCR001$Group<-as.numeric(data_GCCR001$Group)
(bfttest_smell <- ttestBF(formula = Smell_change ~ Group, data = data_GCCR001))[1]
bfttest_smell[1] # Evidence for difference bigger than 0 vs null (no difference)
#bfttest_smell[1] / bfttest_smell[2] # Evidence for difference bigger than 0 vs smaller than zero


#On top of a Theory testing approach, we will quantify the uncertainty about the parameters by sampling from the posteriori through 10000 iteration of the Markov Chain Monte Carlo sampler implemented in the Bayes Factor library:

(bfttest_smell <- ttestBF(formula = Smell_change ~ Group, data = data_GCCR001))
chains = posterior(bfttest_smell, iterations = 1e4)
summary(chains[,'beta (1 - 2)'][chains[, 'beta (1 - 2)'] < 0])
plot(chains[,2])

#PATENCY 
(bfttest_patency <- ttestBF(formula = Nasal_patency_change ~ Group, data = data_GCCR001))[1]
bfttest_patency[1] # Evidence for difference bigger than 0 vs null (no difference)
bfttest_patency[1] / bfttest_patency[2] # Evidence for difference bigger than 0 vs smaller than zero


#On top of a Theory testing approach, we will quantify the uncertainty about the parameters by sampling from the posteriori through 10000 iteration of the Markov Chain Monte Carlo sampler implemented in the Bayes Factor library:

#(bfttest_patency2 <- ttestBF(Nasal_patency_change ~ Group, data = data_GCCR001))
chains = posterior(bfttest_patency, iterations = 1e4)
summary(chains[,'beta (Clinical exam - Lab test)'][chains[, 'beta (Clinical exam - Lab test)'] < 0])


# To test whether smell changes are associated to changes in nasal patency, we will run the following model: 
lmBF(Smell_change ~ Nasal_patency_change * Group + ID, whichRandom ="ID", data = data_GCCR001)

#TASTE
(bfttest_taste <- ttestBF(formula = Taste_change ~ Group, data = data_GCCR001))[1]
bfttest_taste[1] # Evidence for difference bigger than 0 vs null (no difference)
bfttest_taste[1] / bfttest_taste[2] # Evidence for difference bigger than 0 vs smaller than zero


#On top of a Theory testing approach, we will quantify the uncertainty about the parameters by sampling from the posteriori through 10000 iteration of the Markov Chain Monte Carlo sampler implemented in the Bayes Factor library:

#(bfttest2 <- ttestBF(formula = Taste_change ~ Group, data = data_GCCR001))
chains = posterior(bfttest_taste, iterations = 1e4)
summary(chains[,'beta (Clinical exam - Lab test)'][chains[, 'beta (Clinical exam - Lab test)'] < 0])


#CHEMESTHESIS
(bfttest_chemesthesis <- ttestBF(formula = Chemesthesis_change ~ Group, data = data_GCCR001))[1]
bfttest_chemesthesis[1] # Evidence for difference bigger than 0 vs null (no difference)
bfttest_chemesthesis[1] / bfttest_chemesthesis[2] # Evidence for difference bigger than 0 vs smaller than zero


#On top of a Theory testing approach, we will quantify the uncertainty about the parameters by sampling from the posteriori through 10000 iteration of the Markov Chain Monte Carlo sampler implemented in the Bayes Factor library:

#(bfttest2 <- ttestBF(formula = Chemesthesis_change ~ Group, data = data_GCCR001))
chains = posterior(bfttest_chemesthesis, iterations = 1e4)
summary(chains[,'beta (Clinical exam - Lab test)'][chains[, 'beta (Clinical exam - Lab test)'] < 0])

```


# PROPOSAL SHINY APP?

```{r multisensory}
#modalities all together
lmBF((during-before rating) ~ modality*group+ ID, whichRandom="ID", data = data)


colors <- c("#999999", "#E69F00", "#56B4E9")
colors <- colors[as.numeric(data$country)]
scatterplot3d(iris[,6:8], pch = 16, color = colors)
legend("bottom", legend = levels(data$country),
       col =  c("#999999", "#E69F00", "#56B4E9"), pch = 16,
       inset = -0.25, xpd = TRUE, horiz = TRUE)

```
