---
title: "Preliminary analyses GCCR001 April 19th 4 PM"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8)
```

```{r include=FALSE}
##Install/Load Packages
ipak <- function(pkg){
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new.pkg))
    install.packages(new.pkg, dependencies = TRUE, repos = "http://cran.us.r-project.org")
  sapply(pkg, require, character.only = TRUE)
}

packages <- c("psych","ggplot2","tidyverse","scales", "readr", "BayesFactor", "scatterplot3d", "dplyr", "ggstatsplot", "BayesFactor", "summarytools", "PerformanceAnalytics", "nlme", "multcomp", "effects", "LMERConvenienceFunctions", "lme4")

ipak(packages)

```

# Database creation
1. Confirm there are no email duplicates. 

```{r import and prepare database, include = FALSE}
#Database pulled on April 18th 8:34AM ET by Alyssa Joy Bakke, Penn State University
#Accessed on April 19th 4:00AM ET by Valentina Parma for first check
data <- read_csv("/Users/valentinaparma/Documents/GitHub/GCCR001/analysis/paper/Covid_Results_GCCR001.csv")
str(data)


# Include only the relevant variables for the pre-registered GCCR001 analysis, posted at https://osf.io/auhs8/
keeps <- c("Year_of_birth", "Country_of_Residence", "Gender", "Combustible_cigarette_use_no", "E-cigarette_use_no","Symptoms_changes_in_smell", "COVID_diagnosis", "Smell_before_illness", "Smell_during_illness", "Changes_in_smell_i_cannot_smell_at_all_/_smells_smell_less_strong_than_they_did_before", "Changes_in_smell_smells_smell_different_than_they_did_before_(the_quality_of_smell_has_changed)", "Changes_in_smell_i_can_smell_things_that_aren't_there_(for_example_i_smell_burning_when_nothing_is_on_fire)", "Changes_in_smell_sense_of_smell_fluctuates_(comes_and_goes)", "Blocked_nose_before_illness", "Blocked_nose_during_illness", "Taste_before_illness", "Taste_during_illness", "Changes_in_basic_tastes_sweet", "Changes_in_basic_tastes_salty", "Changes_in_basic_tastes_sour", "Changes_in_basic_tastes_bitter", "Changes_in_basic_tastes_savory/umami", "Chemethesis_before_illness","Chemesthesis_during_illness")
data_select<-data[keeps]

# Add ID column
# !!!!!!!!Confirm that there are no duplicate IDs that we can identify via emails 
data_select$ID <- seq.int(nrow(data_select))

# Exclusion criteria. Missing values on smell, taste, chemesthesis ratings before and during
data_select_nomiss<-data_select[!is.na(data_select$Smell_before_illness) & 
                                !is.na(data_select$Smell_during_illness) & 
                                !is.na(data_select$Taste_before_illness) & 
                                !is.na(data_select$Taste_during_illness) & 
                                !is.na(data_select$Chemethesis_before_illness) & 
                                !is.na(data_select$Chemesthesis_during_illness)& 
                                !is.na(data_select$Blocked_nose_before_illness) & 
                                !is.na(data_select$Blocked_nose_during_illness), ]

# Create the change score
data_select_nomiss$Smell_change<-(data_select_nomiss$Smell_during_illness - data_select_nomiss$Smell_before_illness)
data_select_nomiss$Taste_change<-(data_select_nomiss$Taste_during_illness - data_select_nomiss$Taste_before_illness)
data_select_nomiss$Chemesthesis_change<-(data_select_nomiss$Chemesthesis_during_illness - data_select_nomiss$Chemethesis_before_illness)
data_select_nomiss$Nasal_patency_change<-(data_select_nomiss$Blocked_nose_during_illness - data_select_nomiss$Blocked_nose_before_illness)


# Sanity checks. If "Symptoms changing in smell = yes" then one of these must be yes as well 
#"Changes_in_smell_i_cannot_smell_at_all_/_smells_smell_less_strong_than_they_did_before", #"Changes_in_smell_smells_smell_different_than_they_did_before_(the_quality_of_smell_has_changed)", #"Changes_in_smell_i_can_smell_things_that_aren't_there_(for_example_i_smell_burning_when_nothing_is_on_fire)", #"Changes_in_smell_sense_of_smell_fluctuates_(comes_and_goes)"
# If "Symptoms changing in smell = yes" then the smell change score must be above +/-5

# Simplify the visualization of variables
data_select_nomiss$smell_loss<- data_select_nomiss$`Changes_in_smell_i_cannot_smell_at_all_/_smells_smell_less_strong_than_they_did_before`
data_select_nomiss$parosmia<- data_select_nomiss$`Changes_in_smell_smells_smell_different_than_they_did_before_(the_quality_of_smell_has_changed)`
data_select_nomiss$phantosmia<- data_select_nomiss$`Changes_in_smell_i_can_smell_things_that_aren't_there_(for_example_i_smell_burning_when_nothing_is_on_fire)`
data_select_nomiss$smell_fluctuations<- data_select_nomiss$`Changes_in_smell_sense_of_smell_fluctuates_(comes_and_goes)`
  
# Create a new summary sanity check variable
data_select_nomiss<- as.tibble(data_select_nomiss)
data_select_nomiss$Gender<-recode(data_select_nomiss$Gender, `0` = "Female", `1` = "Male", `2` = "Other", `3` = "Prefer not to say")

data_select_nomiss <- mutate(data_select_nomiss, Sanity_check = ifelse(Symptoms_changes_in_smell == 1 && Smell_change >= 5, 1,
                                             ifelse(Symptoms_changes_in_smell == 1 && Smell_change <= 5, 1,
                                             ifelse(Symptoms_changes_in_smell == 1 && data_select_nomiss$smell_loss == 1, 1,
                                             ifelse(Symptoms_changes_in_smell == 1 && data_select_nomiss$parosmia == 1, 1,
                                             ifelse(Symptoms_changes_in_smell == 1 && data_select_nomiss$phantosmia == 1, 1,
                                             ifelse(Symptoms_changes_in_smell == 1 && data_select_nomiss$smell_fluctuations == 1,1,0)))))))                                             
          

# Create COVID-19 diagnostic groups
# COVID_diagnosis = 1 clinical examination
# COVID_diagnosis = 2 & 3 Lab tested

data_select_nomiss<- mutate(data_select_nomiss, Group = ifelse(COVID_diagnosis == 1, "Clinical exam",
                                                        ifelse(COVID_diagnosis == 2 | COVID_diagnosis == 3, "Lab test", "Remove")))

str(data_select_nomiss)

data_select_nomiss$Age <- 2020-data_select_nomiss$Year_of_birth
data_select_nomiss<-as.data.frame(data_select_nomiss)
keeps2 <- c("Age", "Country_of_Residence", "Gender", "Combustible_cigarette_use_no", "E-cigarette_use_no", "COVID_diagnosis", "Smell_change", "Taste_change", "Chemesthesis_change", "Nasal_patency_change", "smell_loss", "parosmia", "phantosmia", "smell_fluctuations","Changes_in_basic_tastes_sweet", "Changes_in_basic_tastes_salty", "Changes_in_basic_tastes_sour", "Changes_in_basic_tastes_bitter", "Changes_in_basic_tastes_savory/umami", "Sanity_check", "Group")
data_select_nomiss2<-data_select_nomiss[keeps2]

# Create final database COVID-19 only
data_GCCR001 <- subset(data_select_nomiss2, Group == "Lab test" | Group == "Clinical exam")
table(data_GCCR001$Country_of_Residence)

data_GCCR001$Group <-factor(data_GCCR001$Group)
data_GCCR001$Gender <-factor(data_GCCR001$Gender)
data_GCCR001$Gender <- relevel(data_GCCR001$Gender, ref = "Male")
data_GCCR001$Gender <- relevel(data_GCCR001$Gender, ref = "Female")
str(data_GCCR001)
data_GCCR001<-data_GCCR001[complete.cases(data_GCCR001),]
```

# Descriptives
```{r echo= FALSE}
summarytools::descr(data_GCCR001)

keeps3<- c("Age", "COVID_diagnosis", "Smell_change", "Taste_change", "Chemesthesis_change", "Nasal_patency_change")
data_GCCR001_descr<-data_GCCR001[keeps3]

# The x axis in each scatter plot represents the column variable, the y axis the row variable. The pairs.panels command will show the pairwise scatter plots of all the variables as well as histograms, locally smoothed regressions, and the Pearson correlation. 
#Note the extreme outlier for the Age = 120, from France ??? 
pairs.panels(data_GCCR001_descr,bg=c("red","blue")[data_GCCR001_descr$COVID_diagnosis],pch=21, main=" Red = COVID-19 clinical exam; Blue = COVID-19 lab tested ")

```

# Age by Group
```{r Age by Group, echo = FALSE, warning=FALSE}
#tiff("GCCR001_PlotAge.tiff", width = 12, height = 5, units = 'in', res = 300)
# for reproducibility
set.seed(123)
# plot
ggstatsplot::grouped_gghistostats(
  data = data_GCCR001,
  x = Age,
  xlab = "Age (in years)",
  type = "robust", # use robust location measure
  grouping.var = Group, # grouping variable
  title.prefix = "Reported diagnosis",
  normal.curve = TRUE, # superimpose a normal distribution curve
  normal.curve.args = list(color = "red", size = 1),
  ggtheme = ggthemes::theme_tufte(),
  ggplot.component = list( # modify the defaults from `ggstatsplot` for each plot
    ggplot2::scale_x_continuous(breaks = seq(0, 100, 10), limits = (c(0, 100)))
  ),
  messages = FALSE,
  #plotgrid.args = list(nrow = 2),
  title.text = "Age in COVID-19 lab tested vs. clinically examined respondents"
)
#dev.off()
```

# Gender by Group
!!!! Order of labels and color of the scale need adjustment
Stats on comparison between groups missing
```{r Gender by Group, echo = FALSE, warning=FALSE}
#tiff("GCCR001_PlotGender.tiff", width = 12, height = 5, units = 'in', res = 300)
# for reproducibility
set.seed(123)

# plot
ggstatsplot::grouped_ggpiestats(
  data = data_GCCR001,
  x = Gender,
  grouping.var = Group, # grouping variable
  title.prefix = "Reported diagnosis", # prefix for the facetted title
  messages = FALSE,
  ggtheme = ggthemes::theme_tufte(),
  title.text = "Gender in COVID-19 lab tested vs. clinically examined respondents"
)
#dev.off()
```


# Smell Change by Group

```{r Smell change by Group, echo = FALSE, warning=FALSE}
data_GCCR001$Group[data_GCCR001$Group == 1] <-"Clinical exam"
data_GCCR001$Group[data_GCCR001$Group == 2] <-"Lab test"
#tiff("GCCR001_PlotSmellChange.tiff", width = 15, height = 5, units = 'in', res = 300)
# for reproducibility
set.seed(123)

# plot
ggstatsplot::grouped_gghistostats(
  data = data_GCCR001,
  x = Smell_change,
  xlab = "Smell change",
  type = "robust", # use robust location measure
  grouping.var = Group, # grouping variable
  title.prefix = "Reported diagnosis",
  normal.curve = TRUE, # superimpose a normal distribution curve
  normal.curve.args = list(color = "red", size = 1),
  ggtheme = ggthemes::theme_tufte(),
  ggplot.component = list( # modify the defaults from `ggstatsplot` for each plot
    ggplot2::scale_x_continuous(breaks = seq(-100, 10, 10), limits = (c(-100, 10))),
    ggplot2::scale_y_continuous(breaks = seq(0, 300, 100), limits = (c(0, 300)))
  ),
  messages = FALSE,
  #plotgrid.args = list(nrow = 2),
  title.text = "Smell change during - before disease in \n COVID-19 lab tested vs. clinically examined respondents"
)
#dev.off()
```


## Characterization - Smell change
```{r smell change characterization, echo = FALSE, warning=FALSE}

#tiff("GCCR001_PlotSmellChange_Characterization.tiff", width = 12, height = 5, units = 'in', res = 300)
# for reproducibility
set.seed(123)

# correlation matrix plot
ggstatsplot::grouped_ggcorrmat(
  data = data_GCCR001,
  cor.vars = c(Smell_change, smell_loss, parosmia, phantosmia, smell_fluctuations, Nasal_patency_change),
  cor.vars.names = c(
    "Smell change", # variable names
    "Loss",
    "Parosmia",
    "Phantosmia",
    "Fluctuations",
    "Nasal patency change"
  ),
  type = "robust", # correlation method
  sig.level = 0.001, # threshold of significance
  p.adjust.method = "holm", # p-value adjustment method for multiple comparisons
  colors = c("#B2182B", "white", "#4D4D4D", "red"),
  grouping.var = Group, # grouping variable
  k = 2, # number of digits after decimal point
  title.prefix = "Reported diagnosis",
  messages = FALSE,
  plotgrid.args = list(ncol = 2))
 # dev.off()
```  
  

# Nasal Patency Change by Group
```{r nasal_patency_change_by_Group, echo = FALSE, warning=FALSE}
data_GCCR001$Group[data_GCCR001$Group == 1] <-"Clinical exam"
data_GCCR001$Group[data_GCCR001$Group == 2] <-"Lab test"
#tiff("GCCR001_PlotNasalPatencyChange.tiff", width = 15, height = 5, units = 'in', res = 300)
# for reproducibility
set.seed(123)

# plot
ggstatsplot::grouped_gghistostats(
  data = data_GCCR001,
  x = Nasal_patency_change,
  xlab = "Nasal patency change",
  type = "robust", # use robust location measure
  grouping.var = Group, # grouping variable
  title.prefix = "Reported diagnosis",
  normal.curve = TRUE, # superimpose a normal distribution curve
  normal.curve.args = list(color = "red", size = 1),
  ggtheme = ggthemes::theme_tufte(),
  ggplot.component = list( # modify the defaults from `ggstatsplot` for each plot
    ggplot2::scale_x_continuous(breaks = seq(-100, 100, 10), limits = (c(-100, 100))),
    ggplot2::scale_y_continuous(breaks = seq(0, 800, 100), limits = (c(0, 800)))
  ),
  messages = FALSE,
  #plotgrid.args = list(nrow = 2),
  title.text = "Nasal patency change during - before disease in \n COVID-19 lab tested vs. clinically examined respondents"
)
#dev.off()
```
 
# Taste Change by Group
```{r taste change, echo = FALSE, warning=FALSE}
data_GCCR001$Group[data_GCCR001$Group == 1] <-"Clinical exam"
data_GCCR001$Group[data_GCCR001$Group == 2] <-"Lab test"
#tiff("GCCR001_PlotTasteChange.tiff", width = 15, height = 5, units = 'in', res = 300)
# for reproducibility
set.seed(123)

# plot
ggstatsplot::grouped_gghistostats(
  data = data_GCCR001,
  x = Taste_change,
  xlab = "Taste change",
  type = "robust", # use robust location measure
  grouping.var = Group, # grouping variable
  title.prefix = "Reported diagnosis",
  normal.curve = TRUE, # superimpose a normal distribution curve
  normal.curve.args = list(color = "red", size = 1),
  ggtheme = ggthemes::theme_tufte(),
  ggplot.component = list( # modify the defaults from `ggstatsplot` for each plot
    ggplot2::scale_x_continuous(breaks = seq(-100, 10, 10), limits = (c(-100, 10))),
    ggplot2::scale_y_continuous(breaks = seq(0, 300, 100), limits = (c(0, 300)))
  ),
  messages = FALSE,
  #plotgrid.args = list(nrow = 2),
  title.text = "Taste change during - before disease in \n COVID-19 lab tested vs. clinically examined respondents"
)
#dev.off()
```


## Characterization - Taste change
```{r taste change characterization, echo= FALSE, warning=FALSE}
data_GCCR001$Group[data_GCCR001$Group == 1] <-"Clinical exam"
data_GCCR001$Group[data_GCCR001$Group == 2] <-"Lab test"
data_GCCR001$sweet<- data_GCCR001$Changes_in_basic_tastes_sweet
data_GCCR001$salt<- data_GCCR001$Changes_in_basic_tastes_salty
data_GCCR001$sour<- data_GCCR001$Changes_in_basic_tastes_sour
data_GCCR001$bitter<- data_GCCR001$Changes_in_basic_tastes_bitter
data_GCCR001$umami<- data_GCCR001$`Changes_in_basic_tastes_savory/umami`

data_GCCR001$Group<-factor(data_GCCR001$Group)
#tiff("GCCR001_PlotTasteChange_Characterization.tiff", width = 12, height = 5, units = 'in', res = 300)
# for reproducibility
set.seed(123)

# correlation matrix plot
ggstatsplot::grouped_ggcorrmat(
  data = data_GCCR001,
  cor.vars = c(Taste_change, sweet, salt, sour, bitter, umami),
  cor.vars.names = c(
    "Taste change", # variable names
    "Sweet",
    "Salt",
    "Sour",
    "Bitter",
    "Umami"
  ),
  type = "robust", # correlation method
  sig.level = 0.001, # threshold of significance
  p.adjust.method = "holm", # p-value adjustment method for multiple comparisons
  colors = c("#B2182B", "white", "#4D4D4D", "red"),
  grouping.var = Group, # grouping variable
  k = 2, # number of digits after decimal point
  title.prefix = "Reported diagnosis",
  messages = FALSE,
  plotgrid.args = list(ncol = 2))

  #dev.off()
```  


# Chemestesis Change by Group
```{r chemesthesis change, echo = FALSE, warning=FALSE}
data_GCCR001$Group[data_GCCR001$Group == 1] <-"Clinical exam"
data_GCCR001$Group[data_GCCR001$Group == 2] <-"Lab test"
#tiff("GCCR001_PlotChemesthesisChange.tiff", width = 15, height = 5, units = 'in', res = 300)
# for reproducibility
set.seed(123)

# plot
ggstatsplot::grouped_gghistostats(
  data = data_GCCR001,
  x = Chemesthesis_change,
  xlab = "Chemesthesis change",
  type = "robust", # use robust location measure
  grouping.var = Group, # grouping variable
  title.prefix = "Reported diagnosis",
  normal.curve = TRUE, # superimpose a normal distribution curve
  normal.curve.args = list(color = "red", size = 1),
  ggtheme = ggthemes::theme_tufte(),
  ggplot.component = list( # modify the defaults from `ggstatsplot` for each plot
    ggplot2::scale_x_continuous(breaks = seq(-100, 10, 100), limits = (c(-100, 100))),
    ggplot2::scale_y_continuous(breaks = seq(0, 300, 100), limits = (c(0, 300)))
  ),
  messages = FALSE,
  #plotgrid.args = list(nrow = 2),
  title.text = "Chemesthesis change during - before disease in \n COVID-19 lab tested vs. clinically examined respondents"
)
#dev.off()
```

# Bayesian analyses 
## Age
```{r Age BF}
data_GCCR001$Group<-as.numeric(data_GCCR001$Group)
(bfttest_Age <- ttestBF(formula = Age ~ Group, data = data_GCCR001))[1]
#bfttest_smell[1] # Evidence for difference bigger than 0 vs null (no difference)
#bfttest_smell[1] / bfttest_smell[2] # Evidence for difference bigger than 0 vs smaller than zero

#On top of a Theory testing approach, we will quantify the uncertainty about the parameters by sampling from the posteriori through 10000 iteration of the Markov Chain Monte Carlo sampler implemented in the Bayes Factor library:
#(bfttest_smell2 <- ttestBF(formula = Smell_change ~ Group, data = data_GCCR001))
#chains_smell = posterior(bfttest_smell2, iterations = 1e4)
#summary(chains_smell[,'beta (1 - 2)'][chains[, 'beta (1 - 2)'] < 0])
#plot(chains_smell[,2])
```
## Smell change
```{r smell change BF}
(bfttest_smell <- ttestBF(formula = Smell_change ~ Group, data = data_GCCR001))[1]
#bfttest_smell[1] # Evidence for difference bigger than 0 vs null (no difference)
#bfttest_smell[1] / bfttest_smell[2] # Evidence for difference bigger than 0 vs smaller than zero

#On top of a Theory testing approach, we will quantify the uncertainty about the parameters by sampling from the posteriori through 10000 iteration of the Markov Chain Monte Carlo sampler implemented in the Bayes Factor library:
#(bfttest_smell2 <- ttestBF(formula = Smell_change ~ Group, data = data_GCCR001))
#chains_smell = posterior(bfttest_smell2, iterations = 1e4)
#summary(chains_smell[,'beta (1 - 2)'][chains[, 'beta (1 - 2)'] < 0])
#plot(chains_smell[,2])
```

## Nasal patency
```{r nasal patency BF}
#(bfttest_nasalpatency <- ttestBF(formula = Nasal_Patency_change ~ Group, data = data_GCCR001))[1]
#bfttest_nasalpatency[1] # Evidence for difference bigger than 0 vs null (no difference)
#bfttest_nasalpatency[1] / bfttest_nasalpatency[2] # Evidence for difference bigger than 0 vs smaller than zero

#On top of a Theory testing approach, we will quantify the uncertainty about the parameters by sampling from the posteriori through 10000 iteration of the Markov Chain Monte Carlo sampler implemented in the Bayes Factor library:
#(bfttest_nasalpatency2 <- ttestBF(formula = Nasal_Patency_change ~ Group, data = data_GCCR001))
#chains_nasalpatency = posterior(bfttest_nasalpatency2, iterations = 1e4)
#summary(chains_nasalpatency[,'beta (1 - 2)'][chains[, 'beta (1 - 2)'] < 0])
#plot(chains_nasalpatency[,2])
```

## Taste change
```{r taste BF}
(bfttest_taste <- ttestBF(formula = Taste_change ~ Group, data = data_GCCR001))[1]
#bfttest_taste[1] # Evidence for difference bigger than 0 vs null (no difference)
#bfttest_taste[1] / bfttest_taste[2] # Evidence for difference bigger than 0 vs smaller than zero

#On top of a Theory testing approach, we will quantify the uncertainty about the parameters by sampling from the posteriori through 10000 iteration of the Markov Chain Monte Carlo sampler implemented in the Bayes Factor library:
#(bfttest_taste2 <- ttestBF(formula = Taste_change ~ Group, data = data_GCCR001))
#chains_taste = posterior(bfttest_taste2, iterations = 1e4)
#summary(chains_taste[,'beta (1 - 2)'][chains[, 'beta (1 - 2)'] < 0])
#plot(chains_taste[,2])
```

## Chemesthesis change
```{r chemesthesis BF}
(bfttest_chemesthesis <- ttestBF(formula = Chemesthesis_change ~ Group, data = data_GCCR001))[1]
#bfttest_chemesthesis[1] # Evidence for difference bigger than 0 vs null (no difference)
#bfttest_chemesthesis[1] / bfttest_chemesthesis[2] # Evidence for difference bigger than 0 vs smaller than zero

#On top of a Theory testing approach, we will quantify the uncertainty about the parameters by sampling from the posteriori through 10000 iteration of the Markov Chain Monte Carlo sampler implemented in the Bayes Factor library:
#(bfttest_chemesthesis2 <- ttestBF(formula = Chemesthesis_change ~ Group, data = data_GCCR001))
#chains_chemesthesis = posterior(bfttest_chemesthesis2, iterations = 1e4)
#summary(chains_chemesthesis[,'beta (1 - 2)'][chains[, 'beta (1 - 2)'] < 0])
#plot(chains_chemesthesis[,2])
```

## Smell nasal patency effect
!!! Computationally intense with random factor
```{r smell nasal patency, echo=FALSE}
# To test whether smell changes are associated to changes in nasal patency, we will run the following model: 
data_GCCR001$ID <- seq.int(nrow(data_GCCR001))
data_GCCR001$ID<-factor(data_GCCR001$ID)
#lmBF(Smell_change ~ Nasal_patency_change * Group + ID, whichRandom ="ID", data = data_GCCR001)
bfMainEffects = lmBF(Smell_change ~ Nasal_patency_change + Group, data = data_GCCR001)
bfInteraction = lmBF(Smell_change ~ Nasal_patency_change + Group + Nasal_patency_change:Group, data = data_GCCR001)
## Compare the two models
bf = bfInteraction / bfMainEffects
bf

newbf = recompute(bf, iterations = 1e4)
newbf
## Sample from the posterior of the full model
chains = posterior(bfInteraction, iterations = 10000)
## 1:13 are the only "interesting" parameters
plot(chains[,2])

ggplot(data_GCCR001, aes(Nasal_patency_change, Smell_change, color = Group)) +
  geom_point() +
  geom_smooth(method = "lm")+ theme_classic()

```


